{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or, How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-01-23 thesis/intro.tex
configure fonts
polish and add some chapter quotes
finish writing landslide static config opts

\chapter{Introduction}
\inspirationalquote{
\begin{tabular}{p{0.65\textwidth}}
Please hold on. This thesis will be departing shortly for the Landslide terminal, baggage claim, ground transportation, and ticketing.
\end{tabular}}
{Pittsburgh International Airport (paraphrased)}

@@

\begin{itemize}
	\item {\bf Background:} Chapter~\ref{chap:background} will present the requisite background material on concurrent programming, stateless model checking, and the various types of programs targeted by Landslide.
	\item {\bf Landslide:} Chapter~\ref{chap:landslide} explains the design and implementation of Landslide
		%the core of my stateless model checking framework,
		and all the special features it's been equipped with over the years.
	\item {\bf Quicksand:} Chapter~\ref{chap:quicksand} presents the Iterative Deepening framework which more intelligently chooses which state spaces to test, corresponding to contribution 1 above.
	\item {\bf Pebbles:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in CMU's 15-410 class environment using the Pebbles kernel, corresponding to contribution 2 above.
	\item {\bf Pintos:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel, corresponding to contribution 2 above.
	\item {\bf Transactional Memory:} Chapter~\ref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.
	\item {\bf Related Work:} Chapter~\ref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
	\item {\bf Conclusion:} Chapter~\ref{chap:conclusion} provides some thoughts on the future of the field.
\end{itemize}

>>>>

\chapter{Introduction}
\inspirationalquote{
\begin{tabular}{p{0.7\textwidth}}
Please hold on. This thesis will be departing shortly for the Landslide terminal, baggage claim, ground transportation, and ticketing.
\end{tabular}}
{Pittsburgh International Airport (paraphrased)}

@@

\begin{itemize}
	\item {\bf Background:} Chapter~\ref{chap:background} will present the requisite background material on concurrent programming, stateless model checking, and the various types of programs targeted by Landslide.
	\item {\bf Landslide:} Chapter~\ref{chap:landslide} explains the design and implementation of Landslide
		%the core of my stateless model checking framework,
		and all the special features it's been equipped with over the years.
	\item {\bf Quicksand:} Chapter~\ref{chap:quicksand} presents the Iterative Deepening framework which more intelligently chooses which state spaces to test, corresponding to contribution 1 above.
	\item {\bf Education:} Chapter~\ref{chap:education} discusses my evaluation of Landslide
		in CMU's 15-410 class environment using the Pebbles kernel,
		and in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel,
		corresponding to contribution 2 above.
	%\item {\bf Pebbles:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in CMU's 15-410 class environment using the Pebbles kernel, corresponding to contribution 2 above.
	%\item {\bf Pintos:} Chapter~\ref{chap:410} discusses my evaluation of Landslide in the University of Chicago's and Berkeley's OS class environments using the Pintos kernel, corresponding to contribution 2 above.
	\item {\bf Transactional Memory:} Chapter~\ref{chap:tm} presents my extension of Landslide's concurrency model to handle transactional concurrency and the evaluation thereof, corresponding to contribution 3 above.
	\item {\bf Related Work:} Chapter~\ref{chap:relatedwork} honors my neighbours and ancestors in research spirit.
	\item {\bf Conclusion:} Chapter~\ref{chap:conclusion} provides some thoughts on the future of the field.
\end{itemize}

### 2018-01-23 thesis/landslide.tex
finish writing landslide static config opts

\begin{itemize}
	\item {\tt -C}: enable ``control experiment'' mode, i.e., run only 1 instance of Landslide, with all (non-data-race) preemption points enabled in advance
		% TODO: put a section reference here
	\item {\tt -I}: enable Iterative Context Bounding (requires {\tt -C}, although future work may relax this restriction);
		this generally causes bugs to be found faster should they exist, but degrades completion time
		(\sect{\ref{sec:landslide-icb}})
	\item {\tt -0}: enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}}, requires {\tt -C})
	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for Pebbles kernelspace mode)
	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for P2 userspace and Pintos modes)
	\item {\tt -X}: support transactional memory (Chapter~\ref{chap:tm})
		% TODO: make a more specific section reference
	\item {\tt -A}: support multiple abort codes during transaction failure (Chapter~\ref{chap:tm});
		required for testing programs which behave differently under different abort circumstances,
		but impacts the state space size
	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
	\item {\tt -4}: support Pebbles architecture (enabled automatically when {\tt p2-setup.sh} is run)
	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c})
	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (described in detail in {\tt id/option.c})
\end{itemize}

@@

The script {\tt pebsim/landslide} invokes Landslide thus.
It should be run from within the {\tt pebsim/} directory.
When supplied no arguments, it reads configuration options from {\tt pebsim/config.landslide}
(a bash script expected to define certain variables as described in \sect{\ref{sec:landslide-glue}}).
The user may optionally specify a file containing additional config directives
% , such as custom preemption points,
as an argument\footnote{
Quicksand actually supplies two such files as arguments: one ``static'' config file and one ''dynamic'' config file.
The former contains options which require recompiling Landslide (e.g., whether or not to use ICB is controlled by an {\tt \#ifdef} in Landslide's code),
while the latter contains options which Landslide interprets at runtime (e.g., which preemption points to use).
The static options do not change between Landslide instances in a single Quicksand run,
avoiding long Landslide start-up times.
	}.
Such supported options are as follows.

@@

Next, configuration options which affect an {\tt \#ifdef} in Landslide and will trigger a recompile upon changing.
Unless otherwise specified these are boolean flags (1 or 0) and the example value shown indicates the default used if unspecified.

\begin{itemize}
	\item {\tt TEST\_CASE} - configure the name of the test program to run (mandatory; no default)
	\item {\tt VERBOSE=0} - enable more verbose output
	\item {\tt FILTER\_DRS\_BY\_TID=1} - configures whether to use the {\tt TID} parameter of {\tt data\_race} described above.
	\item {\tt ICB=0} - enable Iterative Context Bounding (\sect{\ref{sec:landslide-icb}});
		corresponds to {\tt -I} in \sect{\ref{sec:landslide-quicksand-options}}.
	\item {\tt PREEMPT\_EVERYWHERE=0} - enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}});
		corresponds to {\tt -0} in \sect{\ref{sec:landslide-quicksand-options}}.
	\item {\tt PURE\_HAPPENS\_BEFORE=1} - select Pure Happens-Before (1) or Limited Happens-Before (2) (\sect{\ref{sec:background-hb}});
		corresponds to {\tt -V}/{\tt -H} in \sect{\ref{sec:landslide-quicksand-options}}.
	\item % TODO moar
\end{itemize}

>>>>

\begin{itemize}
	\item {\tt -C}: enable ``control experiment'' mode, i.e., run only 1 instance of Landslide, with all (non-data-race) preemption points enabled in advance
		% TODO: put a section reference here
	\item {\tt -I}: enable Iterative Context Bounding (requires {\tt -C}, although future work may relax this restriction);
		this generally causes bugs to be found faster should they exist, but degrades completion time
		(\sect{\ref{sec:landslide-icb}})
	\item {\tt -0}: enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}}, requires {\tt -C})
	\item {\tt -H}: use Limited Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for Pebbles kernelspace mode)
	\item {\tt -V}: use vector-clock-based Pure Happens-Before for data-race analysis (\sect{\ref{sec:background-hb}})
		(default for P2 userspace and Pintos modes)
	\item {\tt -X}: support transactional memory (Chapter~\ref{chap:tm})
		% TODO: make a more specific section reference
	\item {\tt -A}: support multiple abort codes during transaction failure (Chapter~\ref{chap:tm});
		required for testing programs which behave differently under different abort circumstances,
		but impacts the state space size
	\item {\tt -P}: support Pintos architecture (enabled automatically when {\tt pintos-setup.sh} is run)
	\item {\tt -4}: support Pebbles architecture (enabled automatically when {\tt p2-setup.sh} is run)
		% TODO: put a section reference
	\item {\tt -e ETAFACTOR}: configure heuristic state space ETA deferring factor (described in detail in {\tt id/option.c})
	\item {\tt -E ETATHRESH}: configure heuristic threshold of state space progress for judging ETA stability (described in detail in {\tt id/option.c})
\end{itemize}

@@

The script {\tt pebsim/landslide} invokes Landslide thus.
It should be run from within the {\tt pebsim/} directory.
When supplied no arguments, it reads configuration options from {\tt pebsim/config.landslide}
(a bash script expected to define certain variables as described in \sect{\ref{sec:landslide-glue}}).
The user may optionally specify a file containing additional config directives
% , such as custom preemption points,
as an argument.\footnote{
Quicksand actually supplies two such files as arguments: one ``static'' config file and one ''dynamic'' config file.
The former contains options which require recompiling Landslide (e.g., whether or not to use ICB is controlled by an {\tt \#ifdef} in Landslide's code),
while the latter contains options which Landslide interprets at runtime (e.g., which preemption points to use).
The static options do not change between Landslide instances in a single Quicksand run,
avoiding long Landslide start-up times.
}
Such supported options are as follows.

@@

Next, configuration options which affect an {\tt \#ifdef} in Landslide and will trigger a recompile upon changing.
%These span a wide variety of features, sorted below in subsections by roughly how interesting I think they are.
Unless otherwise specified these are boolean flags (1 or 0) and the example value shown indicates the default used if unspecified.

{\bf Search algorithm options}
\begin{itemize}
	\item {\tt ICB=0} - enable Iterative Context Bounding (\sect{\ref{sec:landslide-icb}});
		corresponds to {\tt -I} in \sect{\ref{sec:landslide-quicksand-options}}.
	\item {\tt PREEMPT\_EVERYWHERE=0} - enable Preempt-Everywhere mode (\sect{\ref{sec:quicksand-eval}});
		corresponds to {\tt -0} in \sect{\ref{sec:landslide-quicksand-options}}.
	\item {\tt EXPLORE\_BACKWARDS=0} - configure whether, at each newly encountered preemption point,
		to allow the current thread to run first then later upon backtracking to preempt (0),
		or to issue preemptions first and then try continuing the current thread later (1).
		0 tends to produce shorter preemption traces while 1 tends to find bugs faster (\cite{landslide} \S{}8.7.1).
		Not compatible with ICB.
\end{itemize}

{\bf Memory analysis options}
\begin{itemize}
	\item {\tt PURE\_HAPPENS\_BEFORE=1} - select Pure Happens-Before (1) or Limited Happens-Before (2) (\sect{\ref{sec:background-hb}});
		corresponds to {\tt -V}/{\tt -H} in \sect{\ref{sec:landslide-quicksand-options}}.
	\item {\tt FILTER\_DRS\_BY\_TID=1} - configures whether to use the {\tt TID} parameter of {\tt data\_race} described above.
	\item {\tt FILTER\_DRS\_BY\_LAST\_CALL=0} - configures whether to use the {\tt LAST\_CALL} parameter of {\tt data\_race} described above.
	\item {\tt ALLOW\_LOCK\_HANDOFF=0} - configures lockset tracking to permit or disallow a lock taken by one thread to be released by another thread.%
		\footnote{If enabled, accesses performed by the second thread before unlocking will not be considered protected by that lock,
			as Landslide cannot infer what prior event abstractly represented the lock's ownership changing,
			leading to spurious data race reports.
			This could be solved in future work with a new annotation.}
	\item {\tt ALLOW\_REENTRANT\_MALLOC\_FREE=0} - allow two threads to be in {\tt malloc}, {\tt free}, or so on simultaneously without declaring it a bug.%
		\footnote{Used in Pintos, where those functions lock/unlock the heap mutex themselves rather than relying on a wrapper function to do so before invoking them.}
\end{itemize}

{\bf Interface options}
\begin{itemize}
	\item {\tt TEST\_CASE} - configure the name of the test program to run (mandatory; no default)
	\item {\tt VERBOSE=0} - enable more verbose output
	\item {\tt BREAK\_ON\_BUG=0} - configure whether to exit the simulator or drop into a debug prompt when a bug is found. Simics only and not compatible with Quicksand.
	\item {\tt DONT\_EXPLORE=0} - if enabled, Landslide will not perform stateless model checking but rather will execute the default thread interleaving then exit (useful for manual inspection of preemption points).
	\item {\tt PRINT\_DATA\_RACES=0} - as it says on the tin (for stand-alone use; will message them to Quicksand regardless).
	\item {\tt TABULAR\_TRACE=1} - configure whether to emit bug reports to the console (0) or to an HTML trace file (1)
\end{itemize}

\subsection{Kernel annotations}

The educational experiments in this thesis focus on projects which students implement on top of provided kernel basecode which Landslide already ``understands''.
Such understanding is conferred via the annotations described in this section.
For P2 and Pintos students I supply these annotations behind the scenes,
but a CMU 15-410 student who wishes to use Landslide on their kernel project shall need to brave forth hereupon.

\subsubsection{config.landslide annotations}

The following annotations are specified in {\tt pebsim/config.landslide} akin to the static configuration options described above.
These specify the names of kernel functions, global variables, default values, and so on
which are required to accurately track the kernel's scheduler state:
{\tt CONTEXT\_SWITCH},
{\tt EXEC},
{\tt FIRST\_TID},
{\tt IDLE\_TID},
{\tt INIT\_TID},
{\tt MEMSET},
{\tt PAGE\_FAULT\_WRAPPER},
{\tt PREEMPT\_ENABLE\_FLAG},
{\tt PREEMPT\_ENABLE\_VALUE},
{\tt READLINE},
{\tt SFREE},
{\tt SHELL\_TID},
{\tt SPURIOUS\_INTERRUPT\_WRAPPER},
{\tt THREAD\_KILLED\_ARG\_VAL},
{\tt THREAD\_KILLED\_FUNC},
{\tt TIMER\_WRAPPER},
{\tt VM\_USER\_COPY},
{\tt VM\_USER\_COPY\_TAIL},
{\tt YIELD}.

The following less self-explanatory secret options also exist:
\begin{itemize}
	\item {\tt PINTOS\_KERNEL=0} - configure Landslide for Pebbles (0) or Pintos (1) kernel architecture. Normally set automatically by the setup scripts.
	\item {\tt TESTING\_USERSPACE=1} - configure Landslide whether to test (i.e., focus preemption points, memory analysis, etc. on) the userspace or kernelspace code.
	\item {\tt CURRENT\_THREAD\_LIVES\_ON\_RQ=0} - Landslide infers the list of runnable threads from the {\tt on\_rq()} and {\tt off\_rq()} annotations (described below).
		Some kernels\footnote{most, actually} remove the current thread from their runqueue,
		such that the abstract set of all runnable threads is actually the runqueue plus the current thread rather than just the runqueue.
		Other kernels\footnote{the author's own student kernel from long ago}
		leave the current thread on the runqueue,
		removing it only when it's descheduling and should actually be considered blocked.
		Set this option to 0 to support the former kernel type or 1 to support the latter.
	\item {\tt PATHOS\_SYSCALL\_IRET\_DISTANCE} - indicate how much stack space is used by the reference kernel's system call wrappers (used for cross-kernel-to-userspace stack traces)
	\item {\tt PDE\_PTE\_POISON} - indicate a poison value used in the page tables to indicate absent VM mappings
	\item {\tt BUG\_ON\_THREADS\_WEDGED=1} - set to 0 to disable deadlock detection but instead let the kernel keep receiving system interrupts when all threads appear blocked.%
		\footnote{once used in the bad old days; now recommended for debugging use only}
	\item {\tt TIMER\_WRAPPER\_DISPATCH} - used to manually indicate a label before the end of the timer interrupt assembly wrapper, in case the {\tt iret} instruction couldn't be found automatically (see {\tt pebsim/definegen.sh}).
\end{itemize}

\subsubsection{In-kernel code annotations}

% TODO tell_landslide
